---
title: Weather Data
format: revealjs
---


```{r}
#| echo: false
#| message: false
#| warning: false
knitr::opts_chunk$set(
  message = FALSE,
  warning = FALSE,
  error = FALSE, 
  collapse = TRUE,
  comment = "",
  fig.height = 5,
  fig.width = 8,
  fig.align = "center",
  cache = FALSE,
  echo=FALSE
)
```

```{r}
#| echo: false
#| eval: false
#pak::pak("hadley/emo")
#pak::pak("mitchelloharawild/icons")
```

```{r setup}
library(leaflet)
library(leaflet.providers)
library(readr)
library(tidyverse)
library(htmlwidgets)
library(plotly)
```

## Weather Stations in Utah (30 years plus)

Clicking on point downloads data

```{r}
ut_stations <- read_csv(here::here("slides/data/ut_stations.csv")) |>
  mutate(id = gsub("GHCND:", "", id), 
         years = round((maxdate-mindate)/365)) |>
  mutate(
    file = sprintf("https://www.ncei.noaa.gov/pub/data/ghcn/daily/by_station/%s.csv.gz", id)
  )

ut_stations |>
  filter(years > 30 ) |>
  leaflet() |> 
  addProviderTiles("OpenStreetMap") |>
  addCircleMarkers(lng = ~lon, lat=~lat, 
                   label = ~paste(name, id, 
                                  years, 
                                  sep=" | "),
                   layerId = ~file,
                   radius = 5) |>
  onRender("
    function(el, x) {
      this.eachLayer(function(layer) {
        if (layer instanceof L.CircleMarker) {
          layer.on('click', function(e) {
            var url = layer.options.layerId;

            var a = document.createElement('a');
            a.href = url;
            a.download = '';
            document.body.appendChild(a);
            a.click();
            document.body.removeChild(a);
          });
        }
      });
    }
  ")

```

## Weather in Blanding

```{r read-data}
#files <- dir("data/stations", pattern="csv.gz",
#             full.names = TRUE )
files <- here::here("slides/data/stations/blanding.csv.gz")
weather <- read_csv(files,
col_names = c(
      "id", "date", "element", "value",
      "m_flag", "q_flag", "s_flag", "obs_time"
    ),
    col_types = cols(
      id       = col_character(),
      date     = col_date(format = "%Y%m%d"),
      element  = col_character(),
      value    = col_double(),
      m_flag   = col_character(),
      q_flag   = col_character(),
      s_flag   = col_character(),
      obs_time = col_character()
    ),
    na = c("", "-9999")
)

weather <- weather |> left_join(ut_stations |> select(id, name))
```

<!-- #https://utahweather.org/almanac/daily-weather-almanac-1847-1995 -->
<!-- # # from readme-by-station -->
<!-- # M-FLAG = 1 character Measurement Flag -->
<!-- # Q-FLAG = 1 character Quality Flag -->
<!-- # S-FLAG = 1 character Source Flag -->
<!-- # OBS-TIME = 4-character time of observation -->

<!-- # The five core elements are: -->
<!-- # -->
<!-- # PRCP = Precipitation (tenths of mm) -->
<!-- # SNOW = Snowfall (mm) -->
<!-- # SNWD = Snow depth (mm) -->
<!-- # TMAX = Maximum temperature (tenths of degrees C) -->
<!-- # TMIN = Minimum temperature (tenths of degrees C) -->


<!-- DAEV = Number of days included in the multiday evaporation total (MDEV)  -->
<!-- DAPR = Number of days included in the multiday precipitation total (MDPR)  -->
<!-- DASF = Number of days included in the multiday snowfall total (MDSF)      -->
<!-- DATN = Number of days included in the multiday minimum temperature (MDTN)  -->
<!-- DATX = Number of days included in the multiday maximum temperature (MDTX)  -->
<!-- DAWM = Number of days included in the multiday wind movement (MDWM)  -->
<!-- DWPR = Number of days with non-zero precipitation included in multiday precipitation total (MDPR)  -->

```{r}
# make the data wide to get an overview of which data is included

ww <- weather |> select(id, name, date, element, value) |>
  filter(element %in% c("TMIN", "TMAX", "PRCP", "SNOW", "SNWD")) |>
  pivot_wider(names_from="element", values_from="value") 
  
# assume that missing values in precipitation are 0s:
ww <- ww |> mutate(
  PRCP = ifelse(is.na(PRCP), 0, PRCP),
  SNOW = ifelse(is.na(SNOW), 0, SNOW),
  SNWD = ifelse(is.na(SNWD), 0, SNWD)
)

ww_long <- ww |> pivot_longer(cols=TMAX:SNWD, names_to = "element", 
                              values_to = "value")
```

Station in `r ww$name[1]` reported daily from `r min(ww$date)` to `r max(ww$date)`
(pre-dates founding of the city)

Observed Measurements

- TMAX = Maximum temperature (tenths of degrees C)
- TMIN = Minimum temperature (tenths of degrees C)
- PRCP = Precipitation (tenths of mm)
- SNOW = Snowfall (mm)
- SNWD = Snow depth (mm)

Source for checking on unusual observations: [Daily Almanac](https://utahweather.org/almanac/daily-weather-almanac-1847-1995)

## Daily Precipitation

Average daily precipitation is `r round(mean(ww$PRCP, na.rm= TRUE),1)/254` inch, but it rained only on `r sum(ww$PRCP > 0)` out of `r nrow(ww)` recorded days.

```{r}
gg <- ww_long |> 
  filter(element == "PRCP") |>
  mutate(year = year(date)) |>
  ggplot(aes(x = year, y = value/254)) + 
  geom_point(aes(label = date)) +
  facet_wrap(~month(date, label=TRUE)) +
  geom_smooth(method="lm", aes(group = name)) + 
  theme(legend.position = "bottom") + 
  ggtitle("Daily Precipitation") + ylab("Precipitation (in inch)")

ggplotly(gg, height = 540, width = 1000)
```

## Yearly Total Precipitation


```{r}
prcp <- ww_long |> filter(element == "PRCP") |> mutate(
  year = year(date),
  month = month(date, label=TRUE)
) |> group_by(id, name, year, element) |>
  summarize(
    n = sum(!is.na(value)),
    prcp_day = sum(value > 0, na.rm=TRUE),
    extreme = sum(value > 254, na.rm=TRUE),
    value = sum(value, na.rm=TRUE)/n*365,
    .groups = "drop_last"
  )
```

```{r fig.width = 10}
prcp |> ggplot(aes(x = year, weight = value/254)) + geom_bar() + ggtitle("Amount of total precipitation by year") + ylab("sum(value)") + ylab("Precipitation [inch]")
```

Note: totals are rescaled to 365 days in a year (b/c leap years, missing data)

## Yearly Total Precipitation (2)

```{r fig.width = 10}
prcp |> ggplot(aes(x = year, weight = value/254)) + geom_bar(fill="grey") + ggtitle("Amount of total precipitation by year") + ylab("sum(value)") + 
  geom_point(aes(x = year, y = value/254)) +
  geom_smooth(aes(x = year, y = value/254), method="lm") + ylab("Precipitation [inch]")

model <- lm(value/254 ~ year, data = prcp)
```

## Yearly Total Precipitation (3)

```{r fig.width = 10}
prcp |> ggplot(aes(x = year, weight = value/254)) + geom_bar(fill="grey") + ggtitle("Amount of total precipitation by year") + ylab("sum(value)") + 
  geom_point(aes(x = year, y = value/254)) +
  geom_smooth(aes(x = year, y = value/254), method="lm") + ylab("Precipitation [inch]")

model <- lm(value/254 ~ I(year-1905), data = prcp)
```

```{r}
summary(model)
```
## Monthly Temperatures 

```{r}
temps <- ww_long |> mutate(
  year = year(date),
  month = month(date, label=TRUE)
) |> group_by(id, name, year, month, element) |>
  summarize(
    n = sum(!is.na(value)),
    value = (mean(value, na.rm=TRUE)/10 + 32)*9/5,
    .groups = "drop_last"
  )
```






```{r fig.width = 10, fig.height = 6}
temps |> 
    filter(element %in% c("TMIN", "TMAX"), n > 10) |>
    ggplot(aes(x = year, y = value)) + 
    geom_point(aes(colour = element)) +
    facet_wrap(~month) +
    geom_smooth(method="lm", aes(group = element), colour="grey20") + 
    theme(legend.position = "bottom") + 
    ggtitle("Monthly Averages of \nDaily Minimum and Maximum Temperatures") + 
  ylab("Degree F")
```
## Some of the high Highs

```{r}
temps |> filter(element == "TMAX") |> arrange(desc(value))
```

## Model Maximum temperatures

```{r}
model_tmax <- lm(value~I(year-1905), data = temps |> filter(element == "TMAX"))
model_tmin <- lm(value~I(year-1905), data = temps |> filter(element == "TMIN"))
```

```{r}
summary(model_tmax)
```
## Model Minimum temperatures

```{r}
summary(model_tmin)
```

## So many more questions

- What is the first day of snowfall?
- How long does the snow stay on the ground? 
- Did the average snow depth change?

<br><br>

- What questions do you have? 
<br><br>

![end of excursion]()